Space transformation in Multidimensional IRT (Lemmas)

### 2022-08-05 notebooks/Space_transformation_theorem.qmd
Direction cosines complete
Rebase merge conflict addressed. Adds footnote nº 9

---
title: "Space transformation in Multidimensional IRT (Theorem)"
author: "Mario Luzardo, Daniel Morillo"
format:
  docx:
    number-sections: true
    highlight-style: github  
  html: default
callout-collapse: true
editor_options: 
  chunk_output_type: console
---

@@

In $`r LATENT_SPACE`$ with the standard basis we have the usual scalar product.
In $`r LATENT_SPACE`$ with basis $`r LS_BASIS`$ we define an inner product
$`r INNER_PROD_BASIS`$ such that the norm of $`r TRAIT_VECTOR`$ is invariant.

@@

[^3]: It is ok to use square brackets here? I'm using it instead of parentheses
    because I had already written this (with a function), but if it's not
    correct I change it to parentheses.

>>>>

---
title: "Space transformation in Multidimensional IRT (Lemmas)"
author: "Mario Luzardo, Daniel Morillo"
format:
  docx:
    number-sections: true
    highlight-style: github  
callout-collapse: true
editor_options: 
  chunk_output_type: console
---

@@

In $`r LATENT_SPACE`$ with the standard basis we have the usual scalar product.
In $`r LATENT_SPACE`$ with basis $`r LS_BASIS`$ we define an inner product
$`r INNER_PROD_BASIS`$ such that the norm of $`r TRAIT_VECTOR`$ is invariant,
i.e., $`r TRAIT_NORMS_EQ`$. Thus,[^4]

$$
  `r TRAIT_NORM_STD_EQ`.
$$

Therefore we must define the inner product $`r INNER_PROD_BASIS`$ with Gram
matrix $`r INNER_PROD_MAT_EQ`$. That is, $`r INNER_PROD_EQ`$.[^5]

The Gram matrix has as element $`r INNER_PROD_MAT_EL_EQ`$. As
$`r TRANSFORM_MATRIX`$ (and consequently, $`r TRANSFORM_MATRIX_TRANSP`$) is
invertible so $`r INNER_PROD_MATRIX`$ is, and
$`r INNER_PROD_MATRIX_INV_EQ`$.[^6]

Let's consider now the vectors $`r DIR_COS_VEC`$ and $`r DIR_COS_VEC_STD`$, made
up by $`r N_DIMS`$ direction cosines of $`r TRAIT_VECTOR`$ with the basis
vectors of $`r LS_BASIS`$ and $`r LS_STD_BASIS`$, respectively.[^7]

::: {#lem-dircos}
Under the previously defined conditions, for all $`r TRAIT_VEC_IN_LS`$,[^8]

$$
  `r DIR_COS_VEC_EQ`
$$

with

$$
  `r DIAG_MATRIX_INNER_PROD_INV_SQ_EQ`.
$$

Therefore,

$$
  `r TRAIT_VECTOR_POLAR_EQ2`.
$$

::: proof
For every $`r DIM_INDEX`$; $`r DIM_ENUM_EQ`$,

$$
  `r DIR_COSINE_PROOF`,
$$

with $`r INNER_PROD_MATRIX_ROW`$ the $`r DIM_INDEX`$-th row of
$`r INNER_PROD_MATRIX`$.

Therefore, $`r DIR_COS_VEC_EQ2`$ and $`r TRAIT_VECTOR_POLAR_EQ2`$.[^9]
$\blacksquare$
:::
:::

::: {#lem-convert-spaces}
Let's consider another basis $`r NEW_BASIS_EQ`$ in $`r LATENT_SPACE`$ and the
inner product $`r INNER_PROD_NEWBS`$ that keeps the norm invariant.

Let $`r DIR_COS_VEC_NEWBS`$ be the vector of direction cosines of
$`r TRAIT_VECTOR`$ (i.e., the cosines with the elements of $`r NEW_BASIS`$).

Then,

$$
  `r LEMMA_2_EQ`,
$$ {#eq-lemma-2}

where

$$
  `r DIAG_MATRIX_INNER_PROD_NEWBS_INV_SQ_EQ`
$$

and $`r BASIS_CHANGE_NEWBS_DEF`$.

::: proof
Let's call $`r TRAIT_NEWBS_COORDS_DEF`$ and let $`r BASIS_CHANGE_NEWBS_DEF`$ be
the change of basis matrix such that $`r BASIS_CHANGE_NEWBS`$.

We have that $`r BASES_CHANGE_EQ`$.

The inner product $`r INNER_PROD_NEWBS`$ that keeps the norm invariant is
defined by[^10] $`r INNER_PROD_NEWBS_EQ`$, and we define
$`r INNER_PROD_MAT_NEWBS_EQ`$.

By @lem-dircos, for each case we have:

1.  $`r TRAIT_VECTOR_POLAR_STD_EQ`$

2.  $`r TRAIT_VECTOR_POLAR_EQ2`$

3.  $`r TRAIT_VEC_NEWBS_POLAR_EQ`$

Given (1),

$$
  `r TRAIT_COORDS_EQUIV`
$$

$$
  `r TRAIT_COORDS_EQUIV_NEWBS`.
$$

Using (2) and (3) we get[^11]

$$
  `r TRAIT_COORDS_EQUIV_EXPANDED`
$$

$$
  `r TRAIT_COORDS_EQUIV_NEWBS_EXPANDED`.
$$

Therefore,

$$
  `r TRAIT_COORDS_EQUIV_EXPANDED_EQ`,
$$

and thus $`r LEMMA_2_EQ`$.$\blacksquare$
:::
:::

@@

[^3]: Is it ok to use square brackets here? I'm using it instead of parentheses
    because I had already written this (with a function), but if it's not
    correct I change it to parentheses.

[^4]: Mario wrote in the leftmost therm of the equality "the norm of
    $`r TRAIT_VECTOR`$", but it should be "the *squared* norm of
    $`r TRAIT_VECTOR`$".

[^5]: Having already defined $`r TRAIT_COORDS_DEF`$, I have found this notation
    more convenient (consistent with what I had already written, for which I had
    it already coded).

[^6]: I haven't considered it necessary to add the observation, as I think is a
    rather known theorem in matrix algebra (see e.g. [this response in Math
    Stackexchange](https://math.stackexchange.com/a/340234/1082326 "Response to 'Transpose of inverse vs inverse of transpose'")).

[^7]: I'm omitting here (for the moment) the mathematical expressions defining
    the direction cosine vectors.

[^8]: Mario is using $V$ here but I understand he refers to the latent (vector)
    space, which I have denoted it by $`r LATENT_SPACE`$. Is this correct?

[^9]: I don't quite understand the whole proof, particularly, the equivalence
    $`r latex_eq(latex(latex_transp(STD_BASIS_VECTOR_ANY), INNER_PROD_MATRIX), INNER_PROD_MATRIX_ROW)`$,
    and the step from $`r DIR_COS_ANY`$ to $`r DIR_COS_VEC`$. This step is
    similar to my step from Equation 6 to Equation 7 in the draft, which I had
    the feeling was missing some intermediate steps.

[^10]: Again I'm using here the notation I have used in the draft and before
    (see also footnote 5).

[^11]: I have skipped some basic matrix algebra steps here.
